Appendix 1: Evaluation of the three species in Figure 1

### 2015-08-05 Harris_2015/manuscript-materials/draft.Rmd
submitted to Ecology

\setlength{\parskip}{0pt}

@@

**Abstract:** Estimating species interactions from observational data is one of
the most controversial tasks in community ecology. One difficulty is that a
single pairwise interaction can ripple through an ecological network and produce
surprising indirect consequences. For example, two competing species would
ordinarily correlate negatively in space, but this effect can be reversed in
the presence of a third species that is capable of outcompeting both of them
when it is present. Here, I apply models from statistical physics, called
Markov networks or Markov random fields, that can predict the direct and
indirect consequences of any possible species interaction matrix. Interactions
in these models can be estimated from observational data via maximum
likelihood. Using simulated landscapes with known pairwise interaction
strengths, I evaluated Markov networks and several existing approaches. The
Markov networks consistently outperformed other methods, correctly isolating
direct interactions between species pairs even when indirect interactions or
abiotic environmental effects largely overpowered them. Two computationally
efficient approximations, based on linear and generalized linear models, also
performed well. Indirect effects reliably caused a common null modeling approach
to produce incorrect inferences, however.

@@

\setlength{\parindent}{1em}
\setlength{\parskip}{3pt}

If nontrophic species interactions, such as competition, are important drivers
of community assembly, then ecologists might expect to see their influence in
species composition data [@macarthur_population_1958; @diamond_island_1975].
Despite decades of work and several major controversies, however
[@lewin_santa_1983; @strong_ecological_1984; @gotelli_swap_2003;
@connor_checkered_2013], existing methods for detecting competition's effects
on community structure are unreliable [@gotelli_empirical_2009]. In particular,
species' effects on one another can become lost in the complex web of direct and
indirect interactions in real assemblages. For example, the competitive
interaction between the two shrub species in Figure 1A can become obscured by
their shared tendency to occur in unshaded areas (Figure 1B). While ecologists
have long known that indirect effects can overwhelm direct ones at the landscape
level [@dodson_complementary_1970; @levine_competitive_1976], the vast majority of our methods for drawing
inferenes from observational data do not control for these effects
[e.g. @diamond_island_1975; @strong_ecological_1984; @gotelli_empirical_2009;
@veech_probabilistic_2013; @pollock_understanding_2014]. To the extent that
indirect interactions like those in Figure 1 are generally important
[@dodson_complementary_1970], existing methods will thus not generally provide
much evidence regarding species' direct effects on one another. The goal of this
paper is to resolve this long-standing problem.

While competition doesn't reliably leave the expected pattern in community
structure (low co-occurrence rates at the landscape level), it nevertheless
does leave a signal in the data (Figure 1C). Specifically, *among shaded sites*,
there will be a deficit of co-occurrences between the competing shrub species,
and *among unshaded sites*, there will also be such a deficit. More generally,
we can obtain much better estimates of the association between two species from
their conditional relationships (i.e. by controlling for other species in the
network) than we could get from their overall co-occurrence rates. This kind of
precision is difficult to obtain from null models, which begin with the
assumptions that all the pairwise interactions are zero and thus don’t need to
be controlled for.  Nevertheless, null models have dominated this field for more
than three decades [@strong_ecological_1984; @gotelli_empirical_2009].

Following recent work by @azaele_inferring_2010 and @fort_statistical_2013, this
paper shows that Markov networks [undirected graphical models also known as
Markov random fields; @murphy_machine_2012] can provide a framework for
understanding the landscape-level consequences of pairwise species interactions,
and for detecting them from observed presence-absence matrices. Markov networks
have been used in many scientific fields for decades in similar contexts, from
physics [where nearby particles interact magnetically; @cipra_introduction_1987]
to spatial statistics [where adjacent grid cells have correlated values;
@harris_contact_1974; @gelfand_modelling_2005]. While community ecologists
explored some related approaches in the 1980's [@whittam_species_1981],
they used severe approximations that led to unintelligible results
[e.g. "probabilities" greater than one; @gilpin_factors_1982].

@@

#### Markov networks.
Markov networks provide a framework for translating back and forth between the
conditional relationships among species (Figure 1C) and the kinds of species
assemblages that these relationships produce. Here, I show how a set of
conditional relationships can be used to determine the how groups of species
can co-occur. Methods for estimating conditional relationships from data are
discussed in the next section.

@@

\raggedright
\setlength{\parindent}{1em}

@@

#### Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a matrix of ones and zeros
indicating which species are present at which sites). When the number of species
is reasonably small, one can compute exact maximum likelihood estimates for all
of the $\alpha$ and $\beta$ coefficients given a presence-absence matrix by
optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces [@murphy_machine_2012],
ensuring that this procedure will always converge on the global maximum. This
maximum represents the unique combination of $\alpha$ and $\beta$ coefficients
that would be expected to produce exactly the observed co-occurrence frequencies
on average [i.e. it matches the sufficient statistics of the model distribution
to the sufficient statistics of the data; @murphy_machine_2012]. I used the
rosalia package [@harris_rosalia_2015] for the R programming language
[@r_core_team_r_2015] to optimize the Markov network parameters. The package was
named after Santa Rosalia, the patron saint of biodiversity, whose supposedly
miraculous healing powers played an important rhetorical role in the null model
debates of the 1970's and 1980's [@lewin_santa_1983].

@@

#### Recovering species interactions from simulated data.
I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).

<!-- exp(qlogis(.975, scale = 2)) -->

First, I used the rosalia package [@harris_rosalia_2015] to fit Markov newtork
models,  as described above. For the analyses with 20 species, I added a very
weak logistic prior distribution on the $\alpha$ and $\beta$ terms with scale 2
to ensure that the model estimates were always finite. The bias introduced by
this prior should be small: the 95% credible interval on $\beta$ only requires
that one species' effect on the odds of observing a different species to be less
than a factor of 1500 (which is not much of a constraint). The logistic
distribution was chosen because it is convex and has a similar shape to the
Laplace distribution used in LASSO regularization (especially in the tails), but
unlike the Laplace distribution it is differentiable everywhere and does not
force any estimates to be exactly zero. To confirm that this procedure produced
stable estimates, I compared its estimates on 50 bootstrap replicates
(Appendix 4).

I also evaluated five alternative methods from the existing literature (plus a
novel variant of the fifth method). The first two alternative interaction
measures were the sample correlations and the partial correlations between each
pair of species' data vectors on the landscape [@albrecht_spatial_2001;
@faisal_inferring_2010]. In the context of non-Gaussian data, the partial
correlation can be thought of as a computationally efficient approximation to
the full Markov network model [@loh_structure_2013]. This sort of model is very
common for estimating relationships among genes and gene products
[@friedman_sparse_2008]. Because partial correlations are undefined for
landscapes with perfectly-correlated species pairs, I used a regularized
estimate based on James-Stein shrinkage, as implemented in the corpcor package’s
`pcor.shrink` function with the default settings [@schafer_corpcor_2014].

@@

The fourth alternative method, described in @gotelli_empirical_2009, involved
simulating new landscapes from a null model that retains the row and column
sums of the original matrix [@strong_ecological_1984]. I used the $Z$-scores
computed by the Pairs software described in @gotelli_empirical_2009 as my null
model-based estimator of species interactions.

@@

I then evaluated the relationship between each method's estimates and the "true"
interaction strengths among all of the species pairs from the larger simulated
landscapes. This determined which the methods provide a consistent way to know
how strong species interactions are---regardless of which species were present
in a particular data set or how many observations were taken. Because the
different methods mostly describe species interactions on different scales
(e.g. correlations versus $Z$ scores versus regression coefficients), I used
linear regression through the origin to rescale the different estimates produced
by each method so that they had a consistent interpretation. After rescaling
each method’s estimates, I calculated squared errors between the scaled
interaction estimates and “true” interaction values across all the simulated
data sets. These squared errors determined the proportion of variance explained
for different combinations of method and landscape size (compared with a
null model that assumed all interaction strengths to be zero).

@@

Bootstrap resampling indicated that the above ranking of the different methods
was robust (Appendix 3). In particular, the 95% confidence interval of the
bootstrap distribution indicated that the Markov network's overall $R^2$ value
was between 14 and 18 percent higher than the second-most effective
method (generalized linear models) and between 2.12 and 2.38 times higher than
could be achieved by the null model (Pairs).  Bootstrap resampling of a 200-site
landscape also confirmed that the rosalia package's estimates of species'
conditional relationships were reasonably robust to sampling variation
(Appendix 4).

@@

Despite their limitations, Markov networks have enormous potential to improve
ecological understanding. In particular, they are less vulnerable than some
of the most commonly-used methods to mistakenly identifying positive species
interactions between competing species, and can make precise statements about
the conditions whre indirect interactions will overwhelm direct ones. They also
provide a simple answer to the question of how competition should affect a
species' overall prevalence, which was a major flashpoint for the null model
debates in the 1980’s [@roughgarden_competition_1983; @strong_ecological_1984].
Equation 1 can be used to calculate the expected prevalence of a species in the
absence of biotic influences [$\frac{1}{1 + e^{-\alpha}}$; @lee_learning_2012].
Competition's effect on prevalence in a Markov network can then be calculated by
subtracting this value from the observed prevalence (cf Figure 2D). This kind of
insight would have been difficult to obtain without a generative model that
makes predictions about the consequences of species interactions; null models
(which presume *a priori* that interactions do not exist) have no way to make
such predictions.

Markov networks---particularly the Ising model for binary networks---have been
studied in statistical physics for nearly a century [@cipra_introduction_1987],
and the models' properties, capabilities, and limits are well-understood in a
huge range of applications. Using the same framework for species interactions
would thus allow ecologists to tap into an enormous set of existing
discoveries and techniques for dealing with indirect effects, stability, and
alternative stable states. Numerous other extensions are possible: for example,
the states of the interaction network can be modeled as a function of
the local abiotic environment [@lee_learning_2012], which would provide a
rigorous and straightforward approach to the difficult and important task of
incorporating whole networks of biotic interactions into species distribution
models [@kissling_towards_2012; @pollock_understanding_2014], leading to a
better understanding of the interplay between biotic and abiotic effects on
community structure. There are even methods [@whittam_species_1981;
@tjelmeland_markov_1998] that would allow one species to affect the sign or
strength of the relationship between two other species, tipping the balance
between facilitation and exploitation [@bruno_inclusion_2003].

Finally, the results presented here have important implications for
ecologists' continued use of null models for studying species interactions. Null
and neutral models can be useful for clarifying our thinking about the numerical
consequences of species' richness and abundance patterns
[@harris_occupancy_2011; @xiao_strong_2015], but deviations from a particular
null model must be interpreted with care [@roughgarden_competition_1983]. Even
in small networks with three species, it may simply not be possible to implicate
individual species pairs or specific ecological processes like competition by
rejecting a general-purpose null [@gotelli_empirical_2009], especially when the
test statistic is effectively just a correlation coefficient (Figure 3B).
Simultaneous estimation of multiple ecological parameters seems like a much more
promising approach: to the extent that the models' relative performance on real
data sets is similar to the range of results shown in Figure 3A, scientists in
this field could often double their explanatory power by switching from null
models to Markov networks (or increase it substantially by with linear or
generalized linear approximations). Regardless of the methods ecologists
ultimately choose, controlling for indirect effects could clearly improve our
understanding of species' direct effects on one another and on community
structure.

@@

**Figure 3.**
**A.** Proportion of variance in interaction coefficients explained by each
versus number of sampled locations.
**B.** The $Z$-scores produced by the null model ("Pairs") for each pair of
species are very predictable from the correlation between the presence-absence
vectors of those same species and from the number of sites on the landscape.

>>>>

\setlength{\parskip}{3pt}

@@

**Abstract:** Estimating species interactions from observational data is one of
the most controversial tasks in community ecology. One difficulty is that a
single pairwise interaction can ripple through an ecological network and produce
surprising indirect consequences. For example, the negative correlation between
two competing species can be reversed in the presence of a third species that is
capable of outcompeting both of them. Here, I apply models from statistical
physics, called Markov networks or Markov random fields, that can predict
the direct and indirect consequences of any possible species interaction matrix.
Interactions in these models can be estimated from observational data via
maximum likelihood. Using simulated landscapes with known pairwise interaction
strengths, I evaluated Markov networks and six existing approaches. The
Markov networks consistently outperformed other methods, correctly isolating
direct interactions between species pairs even when indirect interactions
largely overpowered them. Two computationally efficient approximations, based on
linear and generalized linear models, also performed well. Indirect effects
reliably caused a common null modeling approach to produce incorrect inferences,
however.

@@

\setlength{\parindent}{1cm}

To the extent that nontrophic species interactions (such as competition) affect
community assembly, ecologists might expect to find signatures of these interactions in
species composition data [@macarthur_population_1958; @diamond_island_1975].
Despite decades of work and several major controversies, however
[@lewin_santa_1983; @strong_ecological_1984; @gotelli_swap_2003;
@connor_checkered_2013], existing methods for detecting competition's effects
on community structure are unreliable [@gotelli_empirical_2009]. In particular,
species' effects on one another can become lost in the complex web of direct and
indirect interactions in real assemblages. For example, the competitive
interaction between the two shrub species in Figure 1A can become obscured by
their shared tendency to occur in unshaded areas (Figure 1B). While ecologists
have long known that indirect effects can overwhelm direct ones at the landscape
level [@dodson_complementary_1970; @levine_competitive_1976], the vast majority of our methods for drawing
inferenes from observational data do not control for these effects
[e.g. @diamond_island_1975; @strong_ecological_1984; @gotelli_empirical_2009;
@veech_probabilistic_2013; @pollock_understanding_2014]. To the extent that
indirect interactions like those in Figure 1 are generally important
[@dodson_complementary_1970], existing methods will thus not generally provide
much evidence regarding species' direct effects on one another. The goal of this
paper is to resolve this long-standing problem.

While competition doesn't reliably reduce co-occurrence rates at the whole-landscape
level (as most of our methods assume), it nevertheless does leave a signal in
the data (Figure 1C). Specifically, after partitioning the data set into shaded
sites and unshaded sites, there will be co-occurrence deficits in each subset
that might not be apparent at the landscape level. More generally,
we can obtain much better estimates of the association between two species from
their conditional relationships (i.e. by controlling for other species in the
network) than we could get from their overall co-occurrence rates. This kind of
precision is difficult to obtain from null models, which begin with the
assumption that all the pairwise interactions are zero and thus don’t need to
be controlled for.  Nevertheless, null models have dominated this field for more
than three decades [@strong_ecological_1984; @gotelli_empirical_2009].

Following recent work by @azaele_inferring_2010 and @fort_statistical_2013, this
paper shows that Markov networks [undirected graphical models also known as
Markov random fields; @murphy_machine_2012] can provide a framework for
understanding the landscape-level consequences of pairwise species interactions,
and for detecting them from observed presence-absence matrices. Markov networks
have been used in many scientific fields in similar contexts for decades, from
physics [where nearby particles interact magnetically; @cipra_introduction_1987]
to spatial statistics [where adjacent grid cells have correlated values;
@harris_contact_1974; @gelfand_modelling_2005]. While community ecologists
explored some related approaches in the 1980's [@whittam_species_1981],
they used severe approximations that led to unintelligible results
[e.g. "probabilities" greater than one; @gilpin_factors_1982].

@@

#### Markov networks.
Markov networks provide a framework for translating back and forth between the
conditional relationships among species (Figure 1C) and the kinds of species
assemblages that these relationships produce. Here, I show how a set of
conditional relationships can be used to determine how groups of species
can co-occur. Methods for estimating conditional relationships from data are
discussed in the next section.

@@

\raggedright
\setlength{\parindent}{1cm}

@@

#### Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a matrix of ones and zeros
indicating which species are present at which sites). When the number of species
is reasonably small, one can compute exact maximum likelihood estimates for all
of the $\alpha$ and $\beta$ coefficients given a presence-absence matrix by
optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces [@murphy_machine_2012],
ensuring that this procedure will always converge on the global maximum. This
maximum represents the unique combination of $\alpha$ and $\beta$ coefficients
that would be expected to produce exactly the observed co-occurrence frequencies
on average [i.e. maximizing the likelihood matches the sufficient statistics of the model distribution
to the sufficient statistics of the data; @murphy_machine_2012]. I used the
rosalia package [@harris_rosalia_2015] for the R programming language
[@r_core_team_r_2015] to optimize the Markov network parameters. The package was
named after Santa Rosalia, the patron saint of biodiversity, whose supposedly
miraculous healing powers played an important rhetorical role in the null model
debates of the 1970's and 1980's [@lewin_santa_1983].

@@

#### Recovering species interactions from simulated data.
<!-- exp(qlogis(.975, scale = 2)) -->

I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package [@harris_rosalia_2015] to fit Markov newtork
models,  as described above. For the analyses with 20 species, I added a very
weak logistic prior distribution on the $\alpha$ and $\beta$ terms with scale 2
to ensure that the model estimates were always finite. The bias introduced by
this prior should be small: the 95% credible interval on $\beta$ only requires
that one species' effect on the odds of observing a different species to be less
than a factor of 1500 (which is not much of a constraint). The logistic
distribution was chosen because it is convex and has a similar shape to the
Laplace distribution used in LASSO regularization (especially in the tails), but
unlike the Laplace distribution it is differentiable everywhere and does not
force any estimates to be exactly zero. To confirm that this procedure produced
stable estimates, I compared its estimates on 50 bootstrap replicates
(Appendix 4).

I also evaluated six alternative methods: five from the existing literature,
plus a novel combination of two of these methods. The first alternative interaction
metric was the sample correlation between species' presence-absence vectors,
which summarizes their marginal association.  Next, I used partial correlations,
which summarize species' conditional relationships [@albrecht_spatial_2001;
@faisal_inferring_2010]. In the context of non-Gaussian data, the partial
correlation can be thought of as a computationally efficient approximation to
the full Markov network model [@loh_structure_2013]. This sort of model is very
common for estimating relationships among genes and gene products
[@friedman_sparse_2008]. Because partial correlations are undefined for
landscapes with perfectly-correlated species pairs, I used a regularized
estimate based on James-Stein shrinkage, as implemented in the corpcor package’s
`pcor.shrink` function with the default settings [@schafer_corpcor_2014].

@@

The next method, described in @gotelli_empirical_2009, involved
simulating new landscapes from a null model that retains the row and column
sums of the original matrix [@strong_ecological_1984]. I used the $Z$-scores
computed by the Pairs software described in @gotelli_empirical_2009 as my null
model-based estimator of species interactions.

@@

I then evaluated the relationship between each method's estimates and the "true"
interaction strengths among all of the species pairs from the larger simulated
landscapes. This determined which of the methods provide a consistent way to know
how strong species interactions are---regardless of which species were present
in a particular data set or how many observations were taken. Because the
different methods mostly describe species interactions on different scales
(e.g. correlations versus $Z$ scores versus regression coefficients), I used
linear regression through the origin to rescale the different estimates produced
by each method so that they had a consistent interpretation. After rescaling
each method’s estimates, I calculated squared errors between the scaled
interaction estimates and “true” interaction values across all the simulated
data sets. These squared errors determined the proportion of variance explained
for different combinations of model type and landscape size (compared with a
null model that assumed all interaction strengths to be zero).

@@

Bootstrap resampling indicated that the above ranking of the different methods
was robust (Appendix 3). In particular, the 95% confidence interval of the
bootstrap distribution indicated that the Markov network's overall $R^2$ value
was between 14 and 18 percent higher than the second-most effective
method (generalized linear models) and between 2.12 and 2.38 times higher than
could be achieved by the null model (Pairs).  Bootstrap resampling of a 200-site
landscape also confirmed that the rosalia package's estimates of species'
conditional relationships were robust to sampling variation for reasonably-sized
landscapes (Appendix 4).

@@

Despite their limitations, Markov networks have enormous potential to improve
ecological understanding. In particular, they are less vulnerable than some
of the most commonly-used methods to mistakenly identifying positive species
interactions between competing species, and can make precise statements about
the conditions where indirect interactions will overwhelm direct ones. They also
provide a simple answer to the question of how competition should affect a
species' overall prevalence, which was a major flashpoint for the null model
debates in the 1980’s [@roughgarden_competition_1983; @strong_ecological_1984].
Equation 1 can be used to calculate the expected prevalence of a species in the
absence of biotic influences [$\frac{e^\alpha}{1 + e^{\alpha}}$; @lee_learning_2012].
Competition's effect on prevalence in a Markov network can then be calculated by
subtracting this value from the observed prevalence (cf Figure 2D). This kind of
insight would have been difficult to obtain without a generative model that
makes predictions about the consequences of species interactions; null models
(which presume *a priori* that interactions do not exist) have no way to make
such predictions.

Markov networks---particularly the Ising model for binary networks---have been
studied for nearly a century [@cipra_introduction_1987], and the models'
properties, capabilities, and limits are well-understood in a
huge range of applications. Using the same framework for species interactions
would thus allow ecologists to tap into an enormous set of existing
discoveries and techniques for dealing with indirect effects, stability, and
alternative stable states. Numerous other extensions are possible: for example,
the states of the interaction network can be modeled as a function of
the local abiotic environment [@lee_learning_2012], which would provide a
rigorous and straightforward approach to the difficult and important task of
incorporating whole networks of biotic interactions into species distribution
models [@kissling_towards_2012; @pollock_understanding_2014], leading to a
better understanding of the interplay between biotic and abiotic effects on
community structure. There are even methods [@whittam_species_1981;
@tjelmeland_markov_1998] that would allow one species to affect the sign or
strength of the relationship between two other species, tipping the balance
between facilitation and exploitation [@bruno_inclusion_2003].

Finally, the results presented here have important implications for
ecologists' continued use of null models for studying species interactions. Null
and neutral models can be useful for clarifying our thinking about the numerical
consequences of species' richness and abundance patterns
[@harris_occupancy_2011; @xiao_strong_2015], but deviations from a particular
null model must be interpreted with care [@roughgarden_competition_1983]. Even
in small networks with three species, it may simply not be possible to implicate
individual species pairs or specific ecological processes like competition by
rejecting a general-purpose null [@gotelli_empirical_2009], especially when the
test statistic is effectively just a correlation coefficient (Figure 3B).
Simultaneous estimation of multiple ecological parameters seems like a much more
promising approach: to the extent that the models' relative performance on real
data sets is similar to the range of results shown in Figure 3A, scientists in
this field could often double their explanatory power by switching from null
models to Markov networks (or increase it substantially with linear or
generalized linear approximations). Regardless of the methods ecologists
ultimately choose, controlling for indirect effects could clearly improve our
understanding of species' direct effects on one another and on community
structure.

@@

**Figure 3.**
**A.** Proportion of variance in interaction coefficients explained by each
method versus number of sampled locations.
**B.** The $Z$-scores produced by the null model ("Pairs") for each pair of
species can be predicted using the correlation between the presence-absence
vectors of those same species and from the number of sites on the landscape.
